Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-29 paper/paper.tex
added missing file

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls  during analysis.   Next,  the  concrete demand  on  a
function body is obtained through a conservative approximation similar
to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
calls to the function.  The result of the analysis is an annotation of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
program  point where  garbage collection  has been  triggered, the
garbage collector
consults a  set of automata  to restrict reachability  during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

@@

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\input{fig_motiv}
Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of  variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage collectors in reclaiming more garbage.  The specific
contributions of this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language and  prove its correctness. The correctness
  proof involves setting up a non-standard semantics as a
  specification of liveness and then proving that the analysis is
  correct with respect to the specification.   
\item Our  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  regarding  whether   to  copy  a  cell   during  garbage  collection
  translates  to the  membership problem  of such  grammars.  In  this
  paper, we  have shown the  undecidability of the  membership problem
  for  the  class  of  grammars  that result  from  our  analysis.  To
  circumvent the  problem, we  over-approximate the  grammars generated
  by the analysis with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  demonstrate the effectiveness of liveness-based garbage collection.
\end{itemize}
To the  best of our
  knowledge, this is the first work on liveness-based garbage
  collection      for      lazy     languages.       All      previous
  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
    asati14lgc, karkare06effectiveness} have involved either
  imperative or  eager functional languages.

@@

The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
 We assume that each program has a distinguished function
\mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
  \mainpgm})\  e_\mainpgm)$ and  the execution  of the  program starts
with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
$\pi$ (not  part of the language  syntax) with the  program point just
before expression $e$.

@@

The   notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
$\heap[\ell := d]$  indicates the updating of a heap  \heap\ at $\ell$
with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
shadowed  by  $\rho'$  and  $\lfloor \rho  \rfloor_X$  represents  the
environment  restricted  to  the  variables in  $X$.  Finally  $FV(s)$
represents the  free variables  in the  application $s$  and $addr(c)$
gives the address of the closure $c$ in the heap.

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data  needs a richer model  than classical liveness---a
model  which  talks  about  liveness of  references.   Using  $\acar$,
$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
liveness of the structure reachable from a variable can be represented
by a set of {\em access paths} i.e.  prefix-closed set of strings from
$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (typically a
variable  reference) and  a heap  \heap, semantically  an access  path
$\alpha$ represents  a reference, denoted $\heap\oxlb\ell,  \alpha\oxrb$, in
the heap  that is  obtained by  starting with  $\ell$ and  chasing the
\CAR\ or  \CDR\ fields  as specified  by the  access path.  
$\heap\oxlb\ell, \alpha\oxrb$  is defined only  if the path followed  in the
heap  is \emph{closure-free}  (does not  cross closures),  else it  is
undefined.

Access paths can be used to  represent liveness. As an example, if $x$
is  a  list  with  liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
\acdr\acdr,  \acdr\acdr\acar\}$,  then  future computations  can  only
refer up  to the  second and  third members of  $x$.  A  {\em liveness
  environment} is a mapping from  variables to access paths, but often
expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},          y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
represents access using $y$ itself and $z \mapsto \{\}$ indicates that
$z$ is dead.  In lazy  languages, liveness environments are associated
with regions of programs instead of program points.

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

  Since a  closure is evaluated  only at evaluation points,  a closure
  variable is attributed  with the same liveness in the  region of the
  program from the  point of creation to  reachable evaluation points.
  This is  also true  of stack  variables, because,  as we  shall see,
  stack variables derive their  liveness from closure variables. Thus,
  there are  two major differences  in our formulation of  liveness of
  lazy   languages  with   liveness  of   eager 
  languages~\cite{asati14lgc}:    (i)  the
  introduction of  closure variables in the  liveness calculation, and
  (ii) a  single liveness value  for each variable that  is applicable
  from its creation point to evaluation points.

@@

The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\pi_1}$   through  $(\length~\px_{\pi_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\pi_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
%% Also  notice  a  consequence  of laziness:  the  entire  expression
%% including the  condition is not  evaluated if  the demand on  it is
%% $\emptyset$.  This results in the empty liveness environment.
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, and  the  resulting  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$ to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.

As noted earlier, $x.\alpha \in  \Lv_y$ specifies the liveness  of the
reference $\heap\oxlb\rho(x), \alpha\oxrb$  only if $\alpha$ corresponds to  a closure-free path
in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
contents starting from $x_{\pi}$ is given by $\Lv_{x_{\pi}}$.

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial rule is  $\sigma_{all}$ and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it the expression after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environment for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
%%     \oxrb = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the similar to the earlier {\sc car-clo}
rule.

@@

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark consists  of programs taken from nofib~\cite{nofib}
and other similar fucntional programming test suites.  All our
benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
Ubuntu 14.04.

@@

The statistics related to liveness analysis and DFA generation are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this 
overhead, we use a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed.

>>>>

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis  has  two parts.   We  first  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls  during analysis.   The  concrete demand  on  a
function body is obtained through a conservative approximation similar
to  0-CFA~\cite{Shivers:1988} that  combines  the demands  on all  the
calls to the function.  The result of the analysis is an annotation of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
where  garbage collection  has been  triggered, the
collector
consults a  set of automata  to restrict reachability  during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

@@

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\input{fig_motiv}
Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of  variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated due to  the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage collectors in reclaiming more garbage.  The specific
contributions of this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{itemize}
As far as we know,  this is the first work on liveness-based garbage
  collection      for      lazy     languages.       All      previous
  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
    asati14lgc, karkare06effectiveness} have involved either
  imperative or  eager functional languages.

@@

We denote the body of a function ${\mathit  f}$   as $e_{\mathit f}$.
 We assume that each program has a distinguished function
\mainpgm,    defined as    $(\DEFINE\      ({\tt
  \mainpgm})\  e_\mainpgm)$,  and  the program begins execution
with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
$\pi$ (not  part of the syntax) with the  program point just
before expression $e$.

@@

The   notation  $[\myvec{x}   \mapsto  \myvec{\ell}]$   represents  an
environment  that  maps  variables  $x_i$ to  locations  $\ell_i$  and
$\heap[\ell := d]$  indicates updation of  \heap\ at $\ell$
with  $d$.   $\rho \oplus  \rho'$  represents  the environment  $\rho$
shadowed  by  $\rho'$  and  $\lfloor \rho  \rfloor_X$  represents  the
environment  restricted  to  the  variables in  $X$.  Finally  $FV(s)$
represents the  free variables  in the  application $s$  and $addr(c)$
gives the address of the closure $c$ in the heap.

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data  needs a richer model  than classical liveness---a
model  which  talks  about  liveness of  references.   Using  $\acar$,
$\acdr$  to  represent access  using  $\CAR$  and $\CDR$  fields,  the
liveness of the structure reachable from a variable can be represented
by a set of {\em access paths} i.e.  prefix-closed set of strings from
$\{\acar,\acdr\}^\ast$.  Given an initial location $\ell$ (typically a
variable  reference) and  a heap  \heap, semantically  an access  path
$\alpha$ represents  a reference, denoted $\heap\llbracket\ell,  \alpha\rrbracket$, in
the heap  that is  obtained by  starting with  $\ell$ and  chasing the
\CAR\ or  \CDR\ fields in the heap as specified  by the  access path.  
$\heap\llbracket\ell, \alpha\rrbracket$  is defined only  if the path followed  in the
heap  is \emph{closure-free}  (does not  cross closures),  else it  is
undefined.

Access paths are used to  represent liveness. As an example, a list
$x$ with 
liveness  $\{\epsilon,  \acar,  \acdr,  \acdr\acar,
\acdr\acdr,  \acdr\acdr\acar\}$ means that future computations    only
refer up  to the  second and  third members of  $x$.  A  {\em liveness
  environment} is a mapping from  variables to access paths, but often
expressed as  a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},          y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $y   \mapsto  \{\epsilon\}$
represents access using $y$ itself and $z \mapsto \{\}$ indicates that
$z$ is dead.  In lazy  languages, liveness environments are associated
with regions of programs instead of program points.

A notion  that generalizes liveness  is {\em demand}.   While liveness
gives the patterns of future uses of a variable, demand represents the
future uses of the value of an expression.  The demand on an expression
$e$   is   again   a   set    of   access   paths---that   subset   of
$\{\acar,\acdr\}^\ast$ which the  context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the expression
$(\LET~x\leftarrow  (\CDR~y)~\IN~   \SRETURN~x)$.   Assume   that  the
context of this expression places  the demand $\lbrace \epsilon, \acar
\rbrace$. Since the  value of the expression is the  value of $x$, the
demand translates to the liveness  $[x \mapsto \lbrace \epsilon, \acar
  \rbrace]$.   Due  to  the  \LET\   definition  which  binds  $x$  to
$(\CDR~y)$, the liveness of $x$  now becomes the demand on $(\CDR~y)$.
This, in  turn, generates  the liveness $\lbrace  y.\epsilon, y.\acdr,
y.\acdr\acar  \rbrace$.   These are  the  set  of $y$-rooted  accesses
required to  explore $\lbrace  \epsilon, \acar  \rbrace$ paths  of the
result of $(\CDR~y)$.
%% As an  analogy with classical  (strong) liveness analysis,  $y$ and
%% $z$ are live at the entry $\pi: x:=y+z$, if and only if $x$ is live
%% at  exit  of $\pi$.   In  our  terminology, the  liveness  $\lbrace
%% \epsilon\rbrace $ of $x$ at the  exit from $\pi$ becomes the demand
%% on  $y+z$,  and  this,  in turn  generates  the  liveness  $\lbrace
%% y.\epsilon, z.\epsilon \rbrace$ at the entry of $\pi$.

@@

  Since a  closure is evaluated  only at evaluation points,  a closure
  variable is attributed  with the same liveness in the  region of the
  program  from the  point of  creation  of the  closure to  reachable
  evaluation points.  This  is also true of  stack variables, because,
  as we shall see, stack  variables derive their liveness from closure
  variables. Thus, there are two  major differences in our formulation
  of   liveness   of   lazy   languages   with   liveness   of   eager
  languages~\cite{asati14lgc}:   (i)  the   introduction  of   closure
  variables in  the liveness calculation,  and (ii) a  single liveness
  value for each  variable that is applicable from  its creation point
  to evaluation points.

@@

The  variable \pz\   returned  at  $\psi_2$ depends  on  \py\ 
  through  the  expression $(+~\py~1)$.   $\py$  in  turn depends  on  the
  closure  variable  $\px_{\mathit{\pi}_1}$   through  $(\length~\px_{\mathit{\pi}_1})$.   We
  denote  this chain  of dependences  as $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,  \py  \leftarrow   (\length~\px_{\mathit{\pi}_1})\rbrack$.   Indeed,  the
  closures  in  the  heap  are   a  runtime  representation  of  these
  dependences.  Since  $\pz$  is  evaluated   at  $\psi_2$  due  to  the
  expression $\SRETURN~\pz$,  the demand made by  the calling context(s)
  of  $f$ places  a demand  on  $\pz$ which  will impart  a liveness  to
  $\px_{\pi_1}$.  Other dependence chains which result in a liveness for
  $\px_{\pi_1}$       are       $\lbrack       \psi_1\!\!\!:\!\!\py       \leftarrow
  (\length~\px_{\pi_1})\rbrack$ and $\lbrack \psi_3\!\!\!:\!\!\!\pw \leftarrow (/~\pu~\py),\,\,
  \py  \leftarrow  (\length~\px_{\pi_1})\rbrack$.   The  liveness  analysis
  described in this section declares the liveness of $\px_{\pi_1}$ to be
  a union of the liveness arising  out of these dependence chains.  To
  be safe, a  garbage collection during evaluation of  $\py$ at $\psi_1$
  has  to  use  this  liveness  to copy  the  heap  from  $\px_{\pi_1}$.
  However,  notice that  if  a garbage  collection  takes place  while
  evaluating $\pz$ at $\psi_2$, it can safely consider only the liveness
  arising  out of  the dependence  chain\linebreak $\lbrack  \psi_2\!\!\!:\!\!\!\pz \leftarrow
  (+~\py~1),\,   \py  \leftarrow   (\length~\x_{\pi_1})\rbrack$.   The   garbage
  collection  scheme described  in Section~\ref{sec:GC-scheme}  uses a
  generalization  of   this  observation   to  dynamically   select  a
  evaluation point specific liveness in order to collect more garbage.

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, thus the  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$ to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.

As noted earlier, $x.\alpha \in  \Lv_y$ specifies the liveness  of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$ corresponds to  a closure-free path
in \heap\   starting from $\rho(x)$.  If this  path is intercepted  by a
closure,  say  $\CAR~{x_{\pi}}$, then  the  liveness  of further  heap
contents starting from $\mathit{x_{\pi}}$ is given by $\Lv_\mathit{x_{\pi}}$.

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial rule is  $\sigma_{all}$ and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it the expression after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environment for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the similar to the earlier {\sc car-clo}
rule.

@@

\section{Experimental Evaluation}
\label{sec:experiments}
  Our benchmark mostly consists  of programs from nofib~\cite{nofib}
suite.  All our
benchmarks were executed on a  machine having 8 core Intel(R) Core(TM)
i7-4770 3.40GHz  CPU ,  8192 KB  L2 cache,  16 GB  RAM running  64 bit
Ubuntu 14.04.

@@

The statistics related to compile  time liveness analysis are shown in
Table~\ref{tab:exp-results}(a). We observe that the analysis requires
a reasonable time except for  \verb@treejoin@ and  \verb@sudoku@. 
The bottleneck in our analysis is the NFA to DFA conversion
(worst-case exponential behaviour). To avoid this compile-time
overhead, we have added a caching mechanism in our simulator that can
reuse the DFA from earlier analysis if the program has not changed..

Table~\ref{tab:exp-results}(b)   compares   the   garbage   collection
statistics  for RGC  and  LGC.  We  report the  number  of garbage collection  events,
average  number of  cells reclaimed  per garbage collection,  average number  of cells
touched  per garbage collection,  total  number of  dragged cells  and  total time  to
perform  all collections.   It is  no  surprise that  number of  cells
reclaimed per  garbage collection is  higher (and  consequently the  number of  garbage collections is
lower)  for LGC  based collection.   LGC  also reduces  the number  of
dragged cells significantly. However, the cost  for LGC is that per garbage collection
execution  time is  more  for  LGC, which  results  in higher  overall
execution time  for garbage  collection, even  with reduced  number of
collections.   However, LGC  run time  is still  competitive (slowdown
within 5X of RGC for worst case), and is better for 3 benchmarks (best
case  2X speedup).   Note  that  \verb@gc_bench@~\cite{gc_bench} is  a
synthetic benchmark  that allocates complete binary  trees of various
sizes.  These binary trees are never used by the program. As a result,
the benchmark highly favours LGC.  We have included this benchmark for
completeness  only  and   the  numbers  for  it   are  not  considered
representative of practical programs.
